---
title: "fig6_d_e_f_g.Rmd"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 3
  pdf_document: 
    toc: true
    toc_depth: 3
    fig_caption: yes        
    includes:  
      in_header: my_header.tex
    number_sections: true
    keep_tex: true
  word_document: default
date: "2023-07-18"
---

```{r setup, include=FALSE}
suppressPackageStartupMessages(library(gplots))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(ggthemes))
suppressPackageStartupMessages(library(DESeq2))
suppressPackageStartupMessages(library(limma))
suppressPackageStartupMessages(library(edgeR))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(grid))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(UpSetR))
suppressPackageStartupMessages(library(pheatmap))
suppressPackageStartupMessages(library(gplots))
suppressPackageStartupMessages(library(kableExtra))
suppressPackageStartupMessages(library(estimate))
suppressPackageStartupMessages(library(reshape))
suppressPackageStartupMessages(library(sva))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(ggsci))
suppressPackageStartupMessages(library(parallel))

suppressPackageStartupMessages(library(RColorBrewer))
knitr::opts_chunk$set(echo = FALSE, cache=TRUE, warning=FALSE, message=FALSE)
```




```{r read_in_tcl}
library(openxlsx)
drug_screen <- read.xlsx("TCL_screen_data_reformatted.xlsx") 
drug_screen <- drug_screen[!is.na(drug_screen$drug),]
row.names(drug_screen) <- drug_screen$drug
## dropcell line 
drug_screen <- drug_screen[,c(-27,-26,-25,-24)]
drug_screen_just_via <- drug_screen[, c(5:23)]

drug_screen_metadata <- read.xlsx("TCL_screen_data_metadata.xlsx")
#remove cell line
drug_screen_metadata <- drug_screen_metadata[c(-20,-21,-22,-23),]
```

```{r pre-process-kegg-db, eval=F}
library(openxlsx)
library(KEGGREST)
library(data.table)

drug_screen <- read.xlsx("TCL_screen_data_reformatted.xlsx") 
drug_screen <- drug_screen[!is.na(drug_screen$drug),]
row.names(drug_screen) <- drug_screen$drug
#drug_screen$target
## dropcell line 
drug_screen <- drug_screen[,c(-27,-26,-25,-24)]
drug_screen_just_via <- drug_screen[, c(5:23)]

drug_screen <- drug_screen[!duplicated(gsub("\\(.*","",row.names(drug_screen))),]
row.names(drug_screen) <- gsub("\\(.*","",row.names(drug_screen))
drug_screen <- drug_screen[which(row.names(drug_screen) !=  ""),]
drug_ids =  lapply(row.names(drug_screen), function(x) {print(x); keggFind("drug", x)})

drug_ids_df <- lapply(1:length(drug_ids), function(x) {
    if(length(drug_ids[[x]]) > 0 ){
    df <- data.frame(id =names(drug_ids[[x]])[1], name = drug_ids[[x]][[1]], num=x )
      return(df)
    }
  }) %>% rbindlist()
  
drug_info <- lapply(1:nrow(drug_ids_df), function(x) {
      id = as.character(drug_ids_df[x,]$id)
            print(x)
      num = drug_ids_df[x,]$num
      x = keggGet(id)
      target = NA
      pathway = NA
      class=NA
      if(!is.null(x[[1]]$TARGET)){
        if( is.atomic(x[[1]]$TARGET) ){
           target = x[[1]]$TARGET
        } else {
        target = x[[1]]$TARGET$TARGET
        target =  gsub("\\[.*","",target)
      target =  gsub("\\(.*","",target)
         target =  gsub(" ","",target)
        if(!is.null(x[[1]]$TARGET$PATHWAY)){
             pathway =   x[[1]]$TARGET$PATHWAY
              pathway =   gsub("\\(.*","",pathway)
        }
        }
      }
      if(!is.null( x[[1]]$CLASS[[1]])){
              class =  x[[1]]$CLASS[[1]]

      }
      df <- data.frame(id = id , target = paste0(target, collapse = ","), pathway=paste0(pathway, collapse = ","), class=class,num = num)
  return(df)
  })  %>% rbindlist()


drug_info <- merge(drug_info, drug_ids_df, by="num")
drug_info$name  <- as.character(drug_info$name)
#drug_info$name <- gsub("\\(.*","",drug_info$name)
drug_info <- drug_info[!duplicated(drug_info$name),]
drug_info$target  <- as.character(drug_info$target)
drug_info$target <- gsub("\\(.*","",drug_info$target)
drug_info$target <- gsub("\\[.*","",drug_info$target)
drug_info$target <- gsub("/",",",drug_info$target)


saveRDS(drug_info, "drug_info.RDS")

drug_screen_just_via_filt  <- drug_screen_just_via[drug_info$num,]
row.names(drug_screen_just_via_filt)  <- drug_info$name
saveRDS(drug_screen_just_via_filt, "drug_screen_just_via_filt.RDS")
```


```{r load-kegg-drug-db}
drug_info <- readRDS("drug_info.RDS")
drug_screen_just_via_filt <- readRDS("drug_screen_just_via_filt.RDS")
```



```{r load-data}
raw = read.csv("drugscreen.gene.counts.txt", row.names=1, check.names = F)
sub_met = read.csv("drugscreen.metadata.txt", row.names=1, check.names = F)
sub_met$TCL.subtype = factor(sub_met$TCL.subtype)
sub_met$source = factor(sub_met$source)
table(row.names(sub_met) == colnames(raw))
```



```{r pca, fig.align='center', fig.cap="Principal component analysis (PCA) reveals stratification of PDXs and primary tumor samples.", results="asis"}
deseq2.coldata <- data.frame(row.names = colnames(raw), sub_met, stringsAsFactors=F)
deseq2.coldata$source <- factor(deseq2.coldata$source)
deseq2.cds <- DESeq2::DESeqDataSetFromMatrix(countData = raw,colData = deseq2.coldata, design = ~1)
deseq2.cds <- estimateSizeFactors(deseq2.cds)
deseq2.rld <- DESeq2::vst(deseq2.cds, blind=TRUE)
deseq2.rld <- vst(deseq2.cds, blind=TRUE)


ntop = nrow(deseq2.rld)
Pvars <- rowVars(assay(deseq2.rld))
select <- order(Pvars, decreasing = TRUE)[seq_len(min(ntop, length(Pvars)))]
PCA <- prcomp(t(assay(deseq2.rld)[select, ]), scale = F)
percentVar <- round(100*PCA$sdev^2/sum(PCA$sdev^2),1)
dataGG = data.frame(PC1 = PCA$x[,1], PC2 = PCA$x[,2],PC3 = PCA$x[,3], sampleName = row.names(colData(deseq2.rld)),colData(deseq2.rld))
qplot(PC1, PC2, data = dataGG, color =TCL.subtype,  size=I(3), main = "Principal component analysis") + labs(x = paste0("PC1: ", round(percentVar[1],4), "% variance"), y = paste0("PC2: ", round(percentVar[2],4), "% variance")) + theme_bw() + theme(legend.position="bottom") +   scale_color_jco() 
```

# Correlation analysis


To investigate the association between expression and drug sensitivity  at gene level, (Pearson) correlation tests were performed to identify genes whose expressions are correlated with the sensitivity to the drugs with respect to the different TCL subtypes. In other words, is the drug response as measured by cell viability related to expression levels of the different TCL subtypes? 

In order to answer to perform correlation, we need two matrices of equal sizes.  The cell viability assay contains replicates for most PDX samples; thus, we will take the average of each replicate to reduce the measurement of any each PDX sample to one.   

There is not a 1:1 mapping of passage time points in the cell viability data to the expression data.  So, we will take the closest RNA-seq passage. Below is a table of the mapping we used:

```{r mapping, echo=FALSE}
## let's merge reps from viability
il69.p3 <- rowMeans(drug_screen_just_via_filt[,c(1,2)])
il69.p11 <- rowMeans(drug_screen_just_via_filt[,c(3,4)])
DN03.p18 <- rowMeans(drug_screen_just_via_filt[,c(5,6)])
il79.p7 <- drug_screen_just_via_filt[,c(7)]
il79.p8 <- rowMeans(drug_screen_just_via_filt[,c(8,9)])
il2.p5 <- rowMeans(drug_screen_just_via_filt[,c(10,11)])
il2.p7 <- rowMeans(drug_screen_just_via_filt[,c(12,13)])
mt05.p6 <-drug_screen_just_via_filt[,c(14)]
mt05.p8 <-rowMeans(drug_screen_just_via_filt[,c(15,16)])
mt05.p9 <- drug_screen_just_via_filt[,c(17)]
il26a.p4 <- rowMeans(drug_screen_just_via_filt[,c(18,19)])

drugs_via_merge <- data.frame(il69.p3=il69.p3,
                              il69.p11=il69.p11,
                              DN03.p18=DN03.p18,
                              il79.p7=il79.p7,
                              il79.p8=il79.p8,
                              il2.p5=il2.p5,
                              il2.p7=il2.p7,
                              mt05.p6=mt05.p6,
                              mt05.p8=mt05.p8,
                              mt05.p9=mt05.p9,
                              il26a.p4=il26a.p4)
df = data.frame("Drug-assay sample"= colnames(drugs_via_merge), "RNA-seq sample" = colData(deseq2.rld)$sampleName, TCL.subtype=colData(deseq2.rld)$TCL.subtype)
kable(df,  "latex", booktabs = T,  row.names=FALSE, caption = 'Summary of samples for correlation analysis. The cell viability for each drug test was correlated with the expression levels each gene across the different TCL subtypes .') 
```





```{r corr-setup, fig.align='center',  fig.width=12, fig.height=12, results="asis",  echo=FALSE}
table(row.names(sub_met) == colnames(raw))
deseq2.coldata <- data.frame(row.names = colnames(raw), sub_met, stringsAsFactors=F)
deseq2.coldata$source <- factor(deseq2.coldata$source)
deseq2.cds <- DESeq2::DESeqDataSetFromMatrix(countData = raw,colData = deseq2.coldata, design = ~1)
deseq2.cds <- estimateSizeFactors(deseq2.cds)
deseq2.rld <- DESeq2::vst(deseq2.cds, blind=TRUE)
deseq2.rld <- vst(deseq2.cds, blind=TRUE)
deseq2.rld.unfilt <- vst(deseq2.cds, blind=TRUE)
keep <- apply(assay(deseq2.rld), 1, function(x) any(x >= 9)) 
deseq2.rld <- deseq2.rld[keep,]


## let's merge reps from viability
il69.p3 <- rowMeans(drug_screen_just_via_filt[,c(1,2)])
il69.p11 <- rowMeans(drug_screen_just_via_filt[,c(3,4)])
DN03.p18 <- rowMeans(drug_screen_just_via_filt[,c(5,6)])
il79.p7 <- drug_screen_just_via_filt[,c(7)]
il79.p8 <- rowMeans(drug_screen_just_via_filt[,c(8,9)])
il2.p5 <- rowMeans(drug_screen_just_via_filt[,c(10,11)])
il2.p7 <- rowMeans(drug_screen_just_via_filt[,c(12,13)])
mt05.p6 <-drug_screen_just_via_filt[,c(14)]
mt05.p8 <-rowMeans(drug_screen_just_via_filt[,c(15,16)])
mt05.p9 <- drug_screen_just_via_filt[,c(17)]
il26a.p4 <- rowMeans(drug_screen_just_via_filt[,c(18,19)])

drugs_via_merge <- data.frame(il69.p3=il69.p3,
                              il69.p11=il69.p11,
                              DN03.p18=DN03.p18,
                              il79.p7=il79.p7,
                              il79.p8=il79.p8,
                              il2.p5=il2.p5,
                              il2.p7=il2.p7,
                              mt05.p6=mt05.p6,
                              mt05.p8=mt05.p8,
                              mt05.p9=mt05.p9,
                              il26a.p4=il26a.p4)

df = data.frame("Drug-assay sample"= colnames(drugs_via_merge), "RNA-seq sample" = colData(deseq2.rld)$sampleName, TCL.subtype=colData(deseq2.rld)$TCL.subtype)

```





```{r corr, fig.align='center',  fig.width=12, fig.height=12, results="asis", echo=FALSE, eval=T}
drug_info <- as.data.frame(drug_info)
drug_info$pathway <- as.character(drug_info$pathway)
drug_info = drug_info[grep("Ruxolitinib", drug_info$name),]

all = lapply( drug_info$num , function(y) {
  num = as.numeric(y)
  
  df <- drug_info[ drug_info[["num"]] == num , ]
  df <- df[,c(2,7,5,3,4)]
  colnames(df)[1] <- "id"
  
  t_df <- t(df)
  
  
  doi <- as.numeric(drugs_via_merge[as.numeric(colnames(t_df)),] )
  
  env<-new.env()
  env$plots1Names<-c()
  plots1 <- lapply( strsplit(as.character(drug_info[ drug_info[["num"]] == num , ]$target), ",")[[1]] , function(x) {
    #  print(x)
    gene = x
    if(gene != "NA"){
      if(length(grep(gene, row.names(deseq2.rld.unfilt))) > 0){
        lapply(  grep(gene, row.names(deseq2.rld.unfilt), value=T), function(x) {
          #   cat("\n\n##", x, "\n\n")
          df = data.frame(doi=doi, rna=as.numeric(assay(deseq2.rld.unfilt[x,])), subtype =  colData(deseq2.rld.unfilt)$TCL.subtype, patient = colData(deseq2.rld.unfilt)$patient,  passage = colData(deseq2.rld.unfilt)$passage, name = colData(deseq2.rld.unfilt)$sampleName)
          row.names(df) <- colnames(deseq2.rld.unfilt)
          df$subtype <- as.character(df$subtype)
          p1 <- ggbarplot(df, x="name", y="rna", fill="subtype", ylab="Expression level", xlab="", sort.by.groups=F, sort.val="none", color = "white", x.text.angle = 90) + scale_fill_jco() +    coord_flip() + facet_grid(subtype~., scales="free") 
          p2 <- ggscatter(df, x="doi", y="rna", color="subtype", xlab="Cell viability", ylab="Expression level", cor.coef=TRUE, size=5, label="patient", legend = "bottom", title=paste0("Target", " : ", x))  + stat_smooth(method=lm, fill = "lightgray")  + scale_color_jco()  
          #grid.arrange(p1, p2, ncol=2)
          #return(grid.arrange(p1, p2, ncol=2))
          env$plots1Names <- c(env$plots1Names, gene)
          return(arrangeGrob(p1, p2, ncol=2))
        })
      }}
  })
  names(plots1) <- env$plots1Names 
  
  
  tmp = do.call(rbind, mclapply(1:nrow(deseq2.rld), function(i) {
    res = cor.test(doi, assay(deseq2.rld[i,])[1,], method = "pearson")
    data.frame(coef=unname(res$estimate), p=res$p.value)
  }))
  
  corResult <- tibble(ID = rownames(deseq2.rld), 
                      symbol =  rownames(deseq2.rld),
                      coef = tmp$coef,
                      p = tmp$p)
  
  corResult <- arrange(corResult, p) %>% mutate(p.adj = p.adjust(p, method="BH"))
  
  pCut = 0.05
  corResult.sig <- filter(corResult, p <= pCut)
  write.xlsx(corResult.sig,paste0(gsub(" ", "",  drug_info[ drug_info[["num"]] == num , ]$name), ".pearsonCorResultsSig0.05.xlsx"))
  
  
  p1 <- gghistogram(corResult, x="coef", color = "#00AFBB", fill="#00AFBB", rug=T, xlab="Correlation coefficient", main = "Histogram of coefficients")
  p2 <- gghistogram(corResult, x="p", color = "#E7B800", fill="#E7B800", rug=T, xlab = "P value", main = "Histogram of P values") + geom_vline(xintercept=0.05)
  p3 <- gghistogram(corResult.sig, x="coef", color = "#00AFBB", fill="#00AFBB", rug=T, xlab="Correlation coefficient", main = "Histogram of coefficients (P < 0.05)")
  p4 <- gghistogram(corResult.sig, x="p", color = "#E7B800", fill="#E7B800", rug=T, xlab = "P value",  main = "Histogram of P values (P < 0.05)") + geom_vline(xintercept=0.05)
  # grid.arrange(p1, p2, p3, p4, ncol=2)
  
  stats_df = data.frame("genes tested for correlation" = nrow(deseq2.rld), "number correlated (P < 0.05)" = nrow(corResult.sig), check.names = F)
  
  
  gmt.kegg.list <- msigdbr::msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:KEGG") %>% dplyr::select(gs_name, gene_symbol) %>% split(x = .$gene_symbol, f = .$gs_name)
  gmt.kegg <- msigdbr::msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:KEGG") %>% dplyr::select(gs_id, gs_name, gene_symbol) 
  
  
  kegg.gs <- EnrichmentBrowser::getGenesets(org="hsa", db="kegg", cache=FALSE)
  kegg.gs.m = reshape2::melt(kegg.gs)
  kegg.gs.m$value <- as.character(kegg.gs.m$value )
  eg <-  clusterProfiler::bitr(kegg.gs.m$value, fromType="ENTREZID", toType="SYMBOL", OrgDb="org.Hs.eg.db") %>% as.data.table
  kegg.gs.df = merge(kegg.gs.m, eg, by.x="value", by.y="ENTREZID")
  colnames(kegg.gs.df) <- c("entrezid", "pathway", "symbol")
  kegg.gs.df$pathwayid <-  gsub("_.*","",kegg.gs.df$pathway)
  kegg.gs.df$pathwayname <-  sub('.*?_',"",kegg.gs.df$pathway) %>% gsub("_", " ", .)
  
  kegg.gs.list <- kegg.gs.df %>% split(x = .$symbol, f = .$pathwayid)
  
  library(gridExtra)
  
  env<-new.env()
  env$stats<-data.frame()
  env$plots2Names<-c()
  if(drug_info[ drug_info[["num"]] == num , ]$pathway != "NA"){
    plots2 = lapply( strsplit(as.character(drug_info[ drug_info[["num"]] == num , ]$pathway), ",")[[1]] , function(x) {
      set = x
      if(length(corResult.sig$symbol[corResult.sig$symbol %in% kegg.gs.list[[set]]]) >=2){
        #  cat('\\pagebreak')
        #   cat("\n\n##", x, "\n\n")
        env$plots2Names <- c(env$plots2Names, x)
        
        stats.sub = data.frame(set=  length(corResult.sig$symbol[corResult.sig$symbol %in% kegg.gs.list[[set]]]))
        colnames(stats.sub) <- paste0("num sig (P < 0.05) in ", set)
        if (nrow(env$stats) > 0) {
          env$stats <- cbind(env$stats, stats.sub)
        } else {
          env$stats <- stats.sub
        }
        
        
        title =paste0(unique(subset(kegg.gs.df, pathwayid == set)$pathwayid), " : ", unique(subset(kegg.gs.df, pathwayid == set)$pathwayname))
        annotation_col = data.frame(subtype = colData(deseq2.rld)$TCL.subtype, viability = doi)
        row.names(annotation_col) <- colnames(assay(deseq2.rld))
        annotation_row = data.frame(coef=corResult.sig$coef[corResult.sig$symbol %in% kegg.gs.list[[set]]], p=corResult.sig$p[corResult.sig$symbol %in% kegg.gs.list[[set]]])
        row.names(annotation_row) <- corResult.sig$symbol[corResult.sig$symbol %in% kegg.gs.list[[set]]]
        hm = pheatmap(assay(deseq2.rld)[corResult.sig$symbol[corResult.sig$symbol %in% kegg.gs.list[[set]]],], scale="row", cluster_cols = T, annotation_col = annotation_col, annotation_row=annotation_row, color = viridis::inferno(100),cellwidth =25, cellheight = 400/nrow(annotation_row), main = title, silent=T)
        hm2 = pheatmap(assay(deseq2.rld)[corResult.sig$symbol[corResult.sig$symbol %in% kegg.gs.list[[set]]],], scale="row", cluster_cols = F, annotation_col = annotation_col, annotation_row=annotation_row, color = viridis::inferno(100),cellwidth =25, cellheight = 400/nrow(annotation_row), main = title, silent=T) 

        return(arrangeGrob(grobs = list(hm[[4]]), ncol=1))
      } else if (length(corResult.sig$symbol[corResult.sig$symbol %in% kegg.gs.list[[set]]]) > 0){
        env$plots2Names <- c(env$plots2Names, x)
        stats.sub = data.frame(set=  length(corResult.sig$symbol[corResult.sig$symbol %in% kegg.gs.list[[set]]]))
        colnames(stats.sub) <- paste0("num sig (P < 0.05) in ", set)
        if (nrow(env$stats) > 0) {
          env$stats <- cbind(env$stats, stats.sub)
        } else {
          env$stats <- stats.sub
        }
        
        goi = corResult.sig$symbol[corResult.sig$symbol %in% kegg.gs.list[[set]]]
        cpm_df <- assay(deseq2.rld[goi,]) %>% reshape2::melt(., id.vars = rownames)
        names(cpm_df) <- c("gene","sample","exprs")
        cpm_df <- merge(cpm_df, colData(deseq2.rld), by.x = "sample", by.y="sampleName")
        p <- ggdotplot(as.data.frame(cpm_df), x="sample", y="exprs", fill="TCL.subtype", color="TCL.subtype", facet.by=c("gene", "TCL.subtype"), scales="free", ylab = "Expression level")
      }
    })  %>% invisible()
  }
  
  names(plots2) <- env$plots2Names 
  
  cat('\\pagebreak')
  cat("\n\n#", drug_info[ drug_info[["num"]] == num , ]$name, "\n\n")
  
  
  if(nrow(env$stats) > 0 ){
    print(kable(rbind(t_df, t(stats_df), (t(env$stats))),  "latex", booktabs = T,  row.names=T, col.names="") %>% kable_styling(latex_options = c("striped")))
  } else {
    print(kable(rbind(t_df, t(stats_df), t(data.frame("num sig in pathways" = 0,  check.names = F))),  "latex", booktabs = T,  row.names=T, col.names="") %>% kable_styling(latex_options = c("striped")))
  }
  
  print(grid.arrange(p1, p2, p3, p4, ncol=2))
  
  lapply(names(plots1), function(x) {
    if(!is.null(plots1[[x]])){
      x = x
      cat("\n\n##", x, "\n\n")
      grid.arrange((arrangeGrob(grobs = list(plots1[[x]][[1]]))))
      cat('\\pagebreak')
    }
  })
  
  lapply(names(plots2), function(x) {
    if(!is.null(plots2[[x]])){
      cat('\\pagebreak')
      grid.arrange((arrangeGrob(grobs = list(plots2[[x]]))))
      cat("\n\n##", x, "\n\n")
    }
  })
  
})

```




# Session Info

```{r session, message=FALSE, warning=FALSE, cache=FALSE,echo=FALSE, fig.width=10, fig.height=5.5, context="data"}
sessionInfo()
```
